Multi-scale assessment of a grassland productivity model

### 2020-12-27 manuscript/manuscript.Rmd
change errors in text
R2 is now NSE
replace RMSE with mean CVMAE (aka F)
include figure caption clarifying lines and numbers

We evaluated each of the 11 models using the coefficient of determination (R\textsuperscript{2}) and root mean square error (RMSE) of the daily fractional cover estimates. R\textsuperscript{2} and RMSE were calculated for each site and then averaged across all sites within the respective scale. There was no cross-validation using out of sample data in the initial fitting as it would have been computationally expensive. Rather, error metrics from these in-sample tests were treated as a best case scenario in what each model parameterization can achieve. From these results we used a threshold to select which models to evaluate further using cross-validation. The threshold value was an R\textsuperscript{2} threshold of 0.65, which is viewed as "acceptable" for time-series models [@ritter2013].

@@

Models which exceed the threshold were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against this left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam G\textsubscript{cc} was held constant at the value obtained in first fitting. The resulting R\textsuperscript{2} and RMSE are the average among all modelled sites using their respective out of sample test.

@@

At the largest scale, where the PhenoGrass model was parameterized using all 89 sites, the model performed poorly with an average R\textsuperscript{2} value among sites of 0.31 (Table 1; Fig. 3). Models built using all sites of a respective vegetation type performed poorly as well, though were slightly better than the all site model (Fig. 3). The best model performance was achieved when models were built using a single vegetation type subset to a single ecoregion. Grasslands within the Great Plains and Eastern Temperate Forests ecoregions were the only instances where R\textsuperscript{2} exceeded the 0.65 threshold, though Grasslands within N.W. Forests came close (R\textsuperscript{2} = 0.64).

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Observed and predicted fCover values of the All Site model using all available data, and the three vegetation type models each using all PhenoCam sites with an ROI in the respective vegetation type. "}
knitr::include_graphics("figs/allsite_veg_errors.png")
```

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Observed and predicted fCover values for seven model iterations where only sites with a specified vegetation type within a single ecoregion were used in model fitting."}
knitr::include_graphics("figs/ecoregion_vegtype_errors.png")
```

In all 11 iterations the PhenoGrass model tended to underestimate the highest fCover values, and to a lesser degree overpredict the lowest values (Figs. 3,4). The best performing iterations (Grasslands in the Great Plains and Eastern Temperate forests) minimized this effect (Fig. 4). The worst performing iteration, Grasslands in N.A. Deserts, had little variation in predicted fCover values, resulting in the lowest R\textsuperscript{2} overall.

The grassland vegetation type, subset to specific ecoregions, predominantly outperformed other iterations of the PhenoGrass model (Table 1). Models built using grasslands within the Eastern Temperate Forest and Great Plains ecoregions had the highest average R\textsuperscript{2} values of 0.82 and  0.69, respectively. Using leave one out cross-validation on these two grassland model iterations resulted in similar errors of 0.79 and 0.67 for the Eastern Temperate Forest and Great Plains, respectively. Though N.W. Forests grasslands had an in-sample R\textsuperscript{2} just below the 0.65 threshold, the cross-validation was well below it (0.52). Grasslands in the North American deserts were not modelled well at any scale and had the lowest R\textsuperscript{2} values in the entire analysis. The observed greenness patterns of these desert grasslands had extremely high variability in their magnitude and timing, with short distinct peaks in greenness and numerous off-peak fluctuations. The fitted model, which minimized the mean CVME among the 5 sites, was not able to reproduce this high variability and instead produced fCover values that were severely constrained to a narrow range (Fig. S1).

Agriculture and shrubland sites were poorly modelled at all scales. Performance of agriculture within the E. Temperate Forest ecoregion (R\textsuperscript{2} = 0.33) improved over the All Agriculture model (R\textsuperscript{2} = 0.18), but decreased in the Great Plains (from 0.24 to 0.18). There was only a single ecoregion with a minimum of five shrubland sites, N.A. Deserts, and it performed only slightly better than the All Shrubland model. Shrublands in N.A. Deserts did not have the high variability seen in desert grasslands.

\begin{table}
\caption{Average site-level coefficient of determination (R\textsuperscript{2}) and root mean square error (RMSE) for each model parameterization. Bold indicates when the R\textsuperscript{2} was greater than the acceptable threshold of 0.65. Values in parentheses represent the average R\textsuperscript{2} in leave-1-out cross validation.}
\begin{tabular}{l|l|r|r|r}
\hline
 & R\textsuperscript{2} & RMSE & Num. Sites & Site Years\\
\hline
All Sites & 0.31 & 0.14 & 89 & 462.5\\
\hline
All Agriculture & 0.24 & 0.19 & 38 & 175.7\\
\hline
All Grasslands & 0.45 & 0.16 & 37 & 205.8\\
\hline
All Shrublands & 0.44 & 0.09 & 14 & 81.0\\
\hline
\textbf{E. Temperate Forests} &  &  &  & \\
\hline
\hspace{1em}Agriculture & 0.33 & 0.12 & 20 & 99.2\\
\hline
\hspace{1em}Grasslands & \textbf{0.82 (0.79)} & 0.07 & 6 & 35.7\\
\hline
\textbf{Great Plains} &  &  &  & \\
\hline
\hspace{1em}Agriculture & 0.18 & 0.19 & 13 & 54.3\\
\hline
\hspace{1em}Grasslands & \textbf{0.69 (0.67)} & 0.13 & 15 & 71.5\\
\hline
\textbf{N. American Deserts} &  &  &  & \\
\hline
\hspace{1em}Grasslands & -0.04 & 0.19 & 5 & 20.3\\
\hline
\hspace{1em}Shrublands & 0.52 & 0.16 & 8 & 39.7\\
\hline
\textbf{N.W. Forests} &  &  &  & \\
\hline
\hspace{1em}Grasslands & 0.64 (0.52) & 0.08 & 7 & 50.7\\
\hline
\end{tabular}
\end{table}

>>>>

We evaluated each of the 11 models using the Nash Sutcliffe coefficient of efficiency (\(NSE\), Eq. 2) as well as the mean coefficient of variation of the mean absolute error (\textit{F}, Eq. 3) of the daily fractional cover estimates. \(NSE\) was calculated for each site and then averaged across all sites within the respective spatial scale (\( \overline{NSE} \), Eq. 1). There was no cross-validation using out of sample data in the initial fitting as it would have been computationally expensive. Rather, error metrics from these in-sample tests were treated as a best case scenario in what each model parameterization can achieve. From these results we used a threshold to select which models to evaluate further using cross-validation. The threshold value was an \(\overline{NSE}\) of 0.65, which is viewed as "acceptable" for time-series models [@ritter2013]. Parameterization was done using differential evolution, a global optimization algorithm, to minimize \textit{F}.

\begin{equation}
\overline{NSE} = \frac{1}{N}\sum_{j=1}^{N}NSE_{j}
\end{equation}

\begin{equation}
NSE_{j} = 1- \frac{\sum_{i=1}^{n}(fCover_{obs} - fCover_{pred})^{2}}{\sum_{i=1}^{n}(fCover_{obs} - \overline{fCover_{obs}})^{2}}
\end{equation}

\begin{equation}
F = \frac{1}{N}\sum_{j=1}^{N}CVMAE_{j}
\end{equation}

\begin{equation}
CVMAE_{j} = \frac{\frac{1}{i}\sum_{i=1}^{n}|fCover_{i,obs} - fCover_{i,pred}|}{\overline{fCover_{obs}}}
\end{equation}

where \textit{N} is the number of sites within the spatial scale evaluated and \textit{n} is the number of daily values at each site, \textit{fCover\textsubscript{i,obs}} and \textit{fCover\textsubscript{i,pred}} are observed and predicted values, respectively. \textit{fCover\textsubscript{pred}} is the average observed \textit{fCover} at each site.

@@

Models which exceed the threshold were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against this left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam G\textsubscript{cc} was held constant at the value obtained in first fitting. The resulting \(\overline{NSE}\) and \textit{F} are from all modelled sites using their respective out of sample test.

@@

At the largest scale, where the PhenoGrass model was parameterized using all 89 sites, the model performed poorly with an \(\overline{NSE}\) value among sites of 0.31 (Table 1; Fig. 3). Models built using all sites of a respective vegetation type performed poorly as well, though were slightly better than the all site model (Fig. 3). The best model performance was achieved when models were built using a single vegetation type subset to a single ecoregion. Grasslands within the Great Plains and Eastern Temperate Forests ecoregions were the only instances where \(\overline{NSE}\) exceeded the 0.65 threshold, though Grasslands within N.W. Forests came close (NSE = 0.64).

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Observed and predicted fCover values of the All Site model using all available data, and the three vegetation type models each using all PhenoCam sites with an ROI in the respective vegetation type. NSE is the average Nash Sutcliffe coefficient of efficiency among sites, and \textit{F} is the mean coefficient of variation of the mean absolute error. The correlation line (blue) represents the overall trend in predicted versus observed values, while the 1:1 line (red) represents a perfect fit."}
knitr::include_graphics("figs/allsite_veg_errors.png")
```

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Observed and predicted fCover values for seven model iterations where only sites with a specified vegetation type within a single ecoregion were used in model fitting. NSE is the average Nash Sutcliffe coefficient of efficiency among sites, and \textit{F} is the mean coefficient of variation of the mean absolute error. The correlation line (blue) represents the overall trend in predicted versus observed values, while the 1:1 line (red) represents a perfect fit."}
knitr::include_graphics("figs/ecoregion_vegtype_errors.png")
```

In all 11 iterations the PhenoGrass model tended to underestimate the highest fCover values, and to a lesser degree overpredict the lowest values (Figs. 3,4). The best performing iterations (Grasslands in the Great Plains and Eastern Temperate forests) minimized this effect (Fig. 4). The worst performing iteration, Grasslands in N.A. Deserts, had little variation in predicted fCover values, resulting in the lowest \(\overline{NSE}\) overall.

The grassland vegetation type, subset to specific ecoregions, predominantly outperformed other iterations of the PhenoGrass model (Table 1). Models built using grasslands within the Eastern Temperate Forest and Great Plains ecoregions had the highest \(\overline{NSE}\) values of 0.82 and  0.69, respectively. Using leave one out cross-validation on these two grassland model iterations resulted in similar errors of 0.79 and 0.67 for the Eastern Temperate Forest and Great Plains, respectively. Though N.W. Forests grasslands had an in-sample \(\overline{NSE}\) just below the 0.65 threshold, the cross-validation was well below it (0.52). Grasslands in the North American deserts were not modelled well at any scale and had the lowest \(\overline{NSE}\) values in the entire analysis. The observed greenness patterns of these desert grasslands had extremely high variability in their magnitude and timing, with short distinct peaks in greenness and numerous off-peak fluctuations. The fitted model, which minimized the mean CVME among the 5 sites, was not able to reproduce this high variability and instead produced fCover values that were severely constrained to a narrow range (Fig. S1).

Agriculture and shrubland sites were poorly modelled at all scales. Performance of agriculture within the E. Temperate Forest ecoregion (NSE = 0.33) improved over the All Agriculture model (NSE = 0.18), but decreased in the Great Plains (from 0.24 to 0.18). There was only a single ecoregion with a minimum of five shrubland sites, N.A. Deserts, and it performed only slightly better than the All Shrubland model. Shrublands in N.A. Deserts did not have the high variability seen in desert grasslands.

\begin{table}
\caption{Average site-level Nash Sutcliffe coefficient of efficiency (\(\overline{NSE}\)) and mean coefficient of variation of the mean absolute error (\textit{F}) for each model parameterization. Bold indicates when the \(\overline{NSE}\) was greater than the acceptable threshold of 0.65. Values in parentheses represent the \(\overline{NSE}\) or \textit{F} in leave-1-out cross validation.}
\begin{tabular}{l|l|r|r|r}
\hline
& \(\overline{NSE}\) & \textit{F} & Num. Sites & Site Years\\
\hline
All Sites & 0.31 & 0.50 & 89 & 462.5\\
\hline
All Agriculture & 0.24 & 0.51 & 38 & 175.7\\
\hline
All Grasslands & 0.45 & 0.44 & 37 & 205.8\\
\hline
All Shrublands & 0.44 & 0.40 & 14 & 81.0\\
\hline
\textbf{E. Temperate Forests} &  &  &  & \\
\hline
\hspace{1em}Agriculture & 0.33 & 0.45 & 20 & 99.2\\
\hline
\hspace{1em}Grasslands & \textbf{0.82 (0.79)} & 0.28 (0.30) & 6 & 35.7\\
\hline
\textbf{Great Plains} &  &  &  & \\
\hline
\hspace{1em}Agriculture & 0.18 & 0.57 & 13 & 54.3\\
\hline
\hspace{1em}Grasslands & \textbf{0.69 (0.67)} & 0.39 (0.40) & 15 & 71.5\\
\hline
\textbf{N. American Deserts} &  &  &  & \\
\hline
\hspace{1em}Grasslands & -0.04 & 0.44 & 5 & 20.3\\
\hline
\hspace{1em}Shrublands & 0.52 & 0.33 & 8 & 39.7\\
\hline
\textbf{N.W. Forests} &  &  &  & \\
\hline
\hspace{1em}Grasslands & 0.64 & 0.34 & 7 & 50.7\\
\hline
\end{tabular}
\end{table}
